---
title: "LSTM-Calibrated SEIR Model for COVID-19 Dynamics"
subtitle: "Utah COVID-19 Case Prediction with Uncertainty Quantification"
author: "Sima Najafzadehkhoei"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
editor: visual
---

## Introduction

This analysis demonstrates an innovative approach to epidemic modeling by combining deep learning with traditional compartmental models. We use a BiLSTM (Bidirectional Long Short-Term Memory) neural network to calibrate parameters for a stochastic SEIR (Susceptible-Exposed-Infected-Recovered) model, then run multiple simulations to quantify prediction uncertainty.

The SEIR model extends the basic SIR framework by including an **Exposed (E)** compartment, representing individuals who are infected but not yet infectious. This better captures the latent period characteristic of COVID-19 and other infectious diseases.

### Workflow Overview

1.  Download recent COVID-19 case data from Utah
2.  Use BiLSTM to estimate SEIR model parameters from 61-day incidence patterns
3.  Run 2,000 stochastic SEIR simulations with calibrated parameters
4.  Extract daily infection counts (E→I transitions) and compute 95% confidence intervals
5.  Compare model predictions against observed data

## Setup

Load required packages for data manipulation, modeling, and visualization:

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(epiworldR)
})
```

## Data Acquisition

### Download Functions

We create helper functions to download and process Utah's COVID-19 surveillance data:

```{r data-helpers}
source("Real_data.R")

# Now you can use it directly
head(utah_covid_data)
summary(utah_covid_data)
```

### Load Recent Case Data

Retrieve the most recent 61 days of daily case counts:

```{r load-data}
get_covid_data <- function(n_days) {
  
  # Use packaged data
  covid_data <- utah_covid_data
  
  # Filter to last n_days
  last_date <- max(covid_data$Date, na.rm = TRUE)
  covid_data <- subset(covid_data, Date > (last_date - n_days)) %>% 
    dplyr::arrange(Date)
  
  stopifnot("Daily.Cases" %in% names(covid_data))
  covid_data
}
n_days <- 61
covid_data <- get_covid_data(n_days)
incidence <- covid_data$Daily.Cases
incidence_vec <- incidence

stopifnot(length(incidence) == n_days)

cat("Date range:", as.character(min(covid_data$Date)), "to", 
    as.character(max(covid_data$Date)), "\n")
cat("Total cases observed:", sum(incidence), "\n")
cat("Mean daily cases:", round(mean(incidence), 1), "\n")
```

The BiLSTM model requires exactly 61 consecutive days of incidence data. This window length captures approximately two months of epidemic dynamics, sufficient to identify transmission patterns while remaining computationally efficient.

## BiLSTM Parameter Calibration

### Model Parameters

The SEIR model requires four key parameters:

-   **Transmission probability (ptran)**: Probability of infection per contact
-   **Contact rate (crate)**: Average number of contacts per person per day
-   **Incubation rate (incub)**: Rate of progression from exposed to infectious (1/incubation period)
-   **Recovery rate (recov)**: Rate of recovery from infection (1/infectious period)
-   **Basic reproduction number (R₀)**: Expected secondary infections per case

### Running Calibration

We define the population size, incubation rate, and recovery rate, then use the BiLSTM to estimate parameters:

```{r calibrate}

# Model configuration
N     <- 5000        # Population size
incub <- 7.5        # Incubation rate (7.5-day incubation period)
recov <- 1 / 7         # Recovery rate (7-day infectious period)
source("calibrate_seir.R")

# LSTM calibration
lstm_predictions <- calibrate_seir(
  incidence_vec = incidence_vec,
  n     = N,
  incub = incub,
  recov = recov
)

print(lstm_predictions)
```

The BiLSTM neural network has learned relationships between incidence time series patterns and underlying transmission dynamics from training data. It processes the 61-day sequence bidirectionally to capture both past and future context when estimating parameters.

The SEIR model explicitly represents the **latent period** between exposure and infectiousness, which is particularly important for COVID-19 where individuals can be infected for several days before becoming contagious.

### Parameter Consistency

The contact rate is recalculated using the fundamental epidemiological relationship:

$$\text{contact rate} = \frac{R_0 \times \text{recovery rate}}{\text{transmission probability}}$$

This ensures internal consistency between the three parameters.

### Initial Conditions

```{r initial-conditions}
init_infected <- incidence[1]
prev <- init_infected / N

cat("Initial infected count:", init_infected, "\n")
cat("Initial prevalence:", round(prev * 100, 4), "%\n")
```

## Stochastic SEIR Simulation

### Building the Model

We construct a contact-network SEIR model using the calibrated parameters:

```{r build-model}
model <- ModelSEIRCONN(
  name              = "Utah LSTM-calibrated SEIR",
  n                 = N,
  prevalence        = prev,
  contact_rate      = lstm_predictions[["crate"]],
  transmission_rate = lstm_predictions[["ptran"]],
  incubation_days   = incub,
  recovery_rate     = recov
)
```

The `ModelSEIRCONN` structure simulates disease transmission through a contact network with an explicit exposed compartment, providing more realistic dynamics than simple SIR models by accounting for the latent period.

### Multiple Simulation Runs

To quantify uncertainty, we run 2,000 independent stochastic realizations:

```{r run-simulations}
saver <- make_saver("transition")

run_multiple(
  model,
  ndays    = n_days,   # Run for n_days
  nsims    = 2000,
  saver    = saver,
  nthreads = 12
)

sim_results <- run_multiple_get_results(
  model,
  nthreads = 12,
  freader = data.table::fread
)
```

Each simulation represents one possible trajectory of the epidemic. By running many simulations, we capture the stochastic variability inherent in transmission processes.

## Extracting Daily Infections

### Transition Matrices

The model tracks all state transitions. We focus on Exposed → Infected transitions, which represent individuals progressing from the latent period to the infectious period (when cases are typically detected):

```{r extract-transitions}
cat("=== Extracting E→I transitions ===\n")

transitions <- sim_results$transition %>%
  filter(from == "Exposed", to == "Infected") %>%
  arrange(sim_num, date)
# transitions <- sim_results$transition %>%
#   filter(
#     (from == "Susceptible" & to == "Exposed") |
#     (from == "Exposed" & to == "Infected")
#   ) %>%
#   arrange(sim_num, date)

cat("Dimensions:", nrow(transitions), "rows ×", ncol(transitions), "columns\n")
head(transitions, 10)
agg_by_date <- transitions %>%
  group_by(sim_num, date) %>%
  summarise(
    counts = sum(counts),
    .groups = "drop"
  )

```

**Note**: In the SEIR model, E→I transitions represent when exposed individuals become infectious and symptomatic. This is the point at which cases are typically detected and reported. If you want to track new infections (S→E transitions) instead, you can modify the filter to `from == "Susceptible", to == "Exposed"`.

### Computing Confidence Intervals

For each day, we calculate quantiles across all 2,000 simulations:

```{r calculate-quantiles}
cat("=== Calculating 95% CI (2.5th and 97.5th percentiles) ===\n")

quantiles_df <- agg_by_date %>%
  group_by(date) %>%
  summarize(
    lower_ci = quantile(counts, 0.025),
    upper_ci = quantile(counts, 0.975),
    median   = quantile(counts, 0.5),
    mean     = mean(counts),
    .groups  = "drop"
  )

head(quantiles_df, 10)
```

The 95% confidence interval represents the range containing 95% of simulation outcomes, providing a measure of prediction uncertainty.

## Visualization

### Merge with Observed Data

```{r prepare-plot-data}

plot_df <- quantiles_df %>%
  mutate(
    Date           = covid_data$Date,
    observed_cases = incidence  # Use actual observed case data
  )

head(plot_df)
```

### Plot Results with Confidence Intervals

```{r plot-results, fig.width=11, fig.height=7}
p_with_ci <- ggplot(plot_df, aes(x = Date)) +
  # 95% confidence interval
  geom_ribbon(
    aes(ymin = lower_ci, ymax = upper_ci),
    fill = "red",
    alpha = 0.3
  ) +
  # Model median
  geom_line(
    aes(y = median, color = "Model median (E→I)"),
    linewidth = 1.2
  ) +
  # Observed cases
  geom_line(
    aes(y = observed_cases, color = "Observed cases"),
    linewidth = 1.4
  ) +
  geom_point(
    aes(y = observed_cases, color = "Observed cases"),
    size = 2
  ) +
  scale_color_manual(
    values = c(
      "Model median (E→I)" = "red",
      "Observed cases"     = "blue"
    )
  ) +
  labs(
    title = "Daily E→I Transitions: Observed vs Model with 95% CI",
    subtitle = paste0(
      "Population: ", format(N, big.mark = ","),
      " | Contact rate: ", round(lstm_predictions[["crate"]], 3),
      " | Trans. prob: ", round(lstm_predictions[["ptran"]], 3),
      " | Incub. rate: ", round(incub, 3),
      " | R₀: ", round(lstm_predictions[["R0"]], 2)
    ),
    x = "Date",
    y = "Daily E→I Transition Counts",
    color = "",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_with_ci)
```

The red shaded region represents the 95% confidence interval from 2,000 simulations. The red line shows the median prediction, while blue points are observed case counts.

## Model Performance

### Goodness-of-Fit Metrics

```{r performance-metrics}
# CI coverage
coverage <- plot_df %>%
  mutate(in_ci = observed_cases >= lower_ci & observed_cases <= upper_ci) %>%
  summarize(
    coverage_rate = mean(in_ci) * 100,
    n_days = n(),
    days_in_ci = sum(in_ci)
  )

# Error metrics
rmse <- sqrt(mean((plot_df$median - plot_df$observed_cases)^2))
mae <- mean(abs(plot_df$median - plot_df$observed_cases))
mape <- mean(abs((plot_df$observed_cases - plot_df$median) / 
                   plot_df$observed_cases)) * 100

cat("=== Model Performance Metrics ===\n\n")
cat("95% CI Coverage:", round(coverage$coverage_rate, 1), "%\n")
cat("  (", coverage$days_in_ci, "out of", coverage$n_days, "days within CI)\n\n")
cat("Root Mean Squared Error (RMSE):", round(rmse, 2), "cases/day\n")
cat("Mean Absolute Error (MAE):", round(mae, 2), "cases/day\n")
cat("Mean Absolute Percentage Error (MAPE):", round(mape, 1), "%\n")
```

### Interpretation

**Coverage**: A well-calibrated model should have approximately 95% of observed values within the 95% confidence interval. Lower coverage suggests underestimated uncertainty; higher coverage suggests overestimated uncertainty.

**RMSE**: Measures average prediction error magnitude, giving more weight to large errors.

**MAE**: Measures average absolute error without emphasizing outliers.

**MAPE**: Percentage error metric that is scale-independent.

## Discussion

### Key Results

The BiLSTM-calibrated SEIR model successfully captured COVID-19 transmission dynamics in Utah:

-   Estimated transmission probability: **`r round(lstm_predictions[["ptran"]], 3)`**
-   Estimated contact rate: **`r round(lstm_predictions[["crate"]], 3)`** contacts/person/day
-   Incubation rate: **`r round(incub, 3)`** per day (7.5-day incubation period)
-   Recovery rate: **`r round(recov, 3)`** per day (7-day infectious period)
-   Basic reproduction number: **R₀ = `r round(lstm_predictions[["R0"]], 2)`**

The model achieved **`r round(coverage$coverage_rate, 1)`%** coverage with a 95% confidence interval and mean absolute error of **`r round(mae, 2)`** cases per day.

### Advantages of This Approach

1.  **Hybrid methodology**: Combines neural network pattern recognition with mechanistic epidemiological modeling
2.  **Interpretable parameters**: Outputs meaningful epidemiological quantities (R₀, contact rates, incubation period)
3.  **Uncertainty quantification**: Stochastic simulations provide prediction intervals
4.  **Real-time capability**: Can be updated as new surveillance data becomes available
5.  **Minimal assumptions**: BiLSTM learns patterns from data rather than requiring detailed knowledge of transmission mechanisms
6.  **Explicit latent period**: SEIR model better represents COVID-19 biology where individuals are infected but not yet contagious

### Advantages of SEIR over SIR

-   **More realistic disease progression**: Explicitly models the latent/incubation period between infection and infectiousness
-   **Better fit for COVID-19**: COVID-19 has a well-documented incubation period of 5-7 days where individuals are infected but not yet symptomatic or highly contagious
-   **Delays transmission dynamics**: The exposed compartment creates a more realistic delay between infection events and subsequent transmissions
-   **Improved forecasting**: Better captures the temporal dynamics of epidemic curves

### Limitations

-   **Homogeneous mixing**: Assumes uniform contact patterns across the population
-   **Constant parameters**: Does not capture time-varying transmission due to interventions or behavioral changes
-   **Closed population**: No births, deaths, or migration
-   **Fixed incubation and recovery periods**: Assumes all individuals progress through compartments at the same average rate
-   **No spatial structure**: Treats the entire state as a single well-mixed population
-   **Simplified exposed compartment**: Does not distinguish between different stages of the latent period or varying infectiousness

### Future Directions

-   Incorporate time-varying parameters to model interventions (lockdowns, mask mandates)
-   Extend to SEIRS model including waning immunity
-   Add age-structured compartments to capture heterogeneous risk
-   Incorporate spatial structure with metapopulation models
-   Model pre-symptomatic transmission in the exposed compartment
-   Include hospitalization and death compartments for severity predictions

## Conclusion

This analysis demonstrates that BiLSTM neural networks can effectively calibrate traditional compartmental epidemic models from observed incidence data. The SEIR framework provides a more biologically realistic representation of COVID-19 transmission than basic SIR models by explicitly accounting for the latent period between infection and infectiousness.

The resulting stochastic SEIR simulations provide both point predictions and uncertainty quantification, offering a powerful tool for public health decision-making. The approach is particularly valuable when detailed knowledge of transmission mechanisms is limited but historical incidence data is available.

By learning patterns from data, the BiLSTM can estimate parameters that would otherwise require extensive epidemiological investigation, while the SEIR structure ensures the model respects fundamental biological constraints about disease progression.
