---
title: "LSTM-Calibrated SEIR Model for COVID-19 Dynamics"
subtitle: "Utah COVID-19 Case Prediction with Uncertainty Quantification"
author: "Sima Najafzadehkhoei"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
editor: visual
---

## Introduction

This analysis demonstrates an innovative approach to epidemic modeling by combining deep learning with traditional compartmental models. We use a BiLSTM (Bidirectional Long Short-Term Memory) neural network to calibrate parameters for a stochastic SEIR (Susceptible-Exposed-Infected-Recovered) model, then run multiple simulations to quantify prediction uncertainty.

The SEIR model extends the basic SIR framework by including an **Exposed (E)** compartment, representing individuals who are infected but not yet infectious. This better captures the latent period characteristic of COVID-19 and other infectious diseases.

### Workflow Overview

1.  Download recent COVID-19 case data from Utah
2.  Use BiLSTM to estimate β, R₀, and incubation period from 61-day incidence patterns
3.  Derive ptran and crate from β using a fixed contact rate assumption
4.  Run 2,000 stochastic SEIR simulations with calibrated parameters
5.  Extract daily infection counts (E→I transitions) and compute 95% confidence intervals
6.  Compare model predictions against observed data

## Setup

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(epiworldR)
})
```

## Data Acquisition

```{r data-helpers}
source("Real_data.R")

head(utah_covid_data)
summary(utah_covid_data)
```

### Load Recent Case Data

```{r load-data}
get_covid_data <- function(n_days) {
  covid_data <- utah_covid_data
  last_date  <- max(covid_data$Date, na.rm = TRUE)
  covid_data <- subset(covid_data, Date > (last_date - n_days)) %>%
    dplyr::arrange(Date)
  stopifnot("Daily.Cases" %in% names(covid_data))
  covid_data
}

n_days        <- 61
covid_data    <- get_covid_data(n_days)
incidence     <- covid_data$Daily.Cases
incidence_vec <- incidence

stopifnot(length(incidence) == n_days)

cat("Date range:", as.character(min(covid_data$Date)), "to",
    as.character(max(covid_data$Date)), "\n")
cat("Total cases observed:", sum(incidence), "\n")
cat("Mean daily cases:", round(mean(incidence), 1), "\n")
```

## BiLSTM Parameter Calibration

### Model Parameters

The BiLSTM model predicts three quantities from the incidence time series:

-   **β (beta)**: Composite transmission rate = ptran × crate
-   **R₀**: Basic reproduction number = β / recov
-   **incub**: Incubation period in days

`ModelSEIRCONN` requires `ptran` and `crate` as separate inputs. Since the model predicts only their product β, we fix `crate` at an epidemiologically plausible value for COVID-19 and derive `ptran = β / crate`.

### Running Calibration

```{r calibrate}
# Fixed parameters
N     <- 5000    # Population size
recov <- 1 / 7   # Recovery rate (7-day infectious period)

# Contact rate assumption for COVID-19 (used to split beta into ptran and crate)
# Typical range for COVID-19: 8–15 contacts/person/day
assumed_crate <- 10

source("calibrate_seir.R")

# LSTM calibration — returns a data.table with columns: beta, R0, incub
lstm_predictions <- calibrate_seir_revin(
  incidence_vec = incidence_vec,
  n             = N,
  recov         = recov
)

print(lstm_predictions)
```

### Derive ptran and crate from β

```{r derive-params}
# Extract model outputs using $ (lstm_predictions is a data.table)
beta  <- lstm_predictions$beta
R0    <- lstm_predictions$R0
incub <- lstm_predictions$incub

# Derive ptran and crate from beta = ptran * crate
# R0 = ptran * crate / recov  →  beta = R0 * recov  (consistency check)
crate <- assumed_crate
ptran <- beta / crate

cat("=== Calibrated SEIR Parameters ===\n")
cat(sprintf("  beta  (model output)         : %.6f\n", beta))
cat(sprintf("  R0    (model output)         : %.4f\n",  R0))
cat(sprintf("  incub (model output)         : %.2f days\n", incub))
cat(sprintf("  recov (fixed)                : %.4f\n",  recov))
cat(sprintf("  crate (fixed assumption)     : %.1f contacts/day\n", crate))
cat(sprintf("  ptran (derived = beta/crate) : %.6f\n", ptran))
```

### Parameter Consistency Check

```{r consistency-check}
# R0 should equal ptran * crate / recov = beta / recov
R0_check <- beta / recov
R0_diff  <- abs(R0 - R0_check)

cat("=== SEIR Consistency Check (R0 = beta / recov) ===\n")
cat(sprintf("  R0 from model  : %.4f\n", R0))
cat(sprintf("  beta / recov   : %.4f\n", R0_check))
cat(sprintf("  Difference     : %.6f  %s\n", R0_diff,
    ifelse(R0_diff < 0.05, "✓ PASS", "✗ FAIL")))
```

### Initial Conditions

```{r initial-conditions}
init_infected <- incidence[1]
prev          <- init_infected / N

cat("Initial infected count:", init_infected, "\n")
cat("Initial prevalence:", round(prev * 100, 4), "%\n")
```

## Stochastic SEIR Simulation

### Building the Model

```{r build-model}
model <- ModelSEIRCONN(
  name              = "Utah LSTM-calibrated SEIR",
  n                 = N,
  prevalence        = prev,
  contact_rate      = crate,
  transmission_rate = ptran,
  incubation_days   = incub,
  recovery_rate     = recov
)
```

### Multiple Simulation Runs

```{r run-simulations}
saver <- make_saver("transition")

run_multiple(
  model,
  ndays    = n_days,
  nsims    = 2000,
  saver    = saver,
  nthreads = 12
)

sim_results <- run_multiple_get_results(
  model,
  nthreads = 12,
  freader  = data.table::fread
)
```

## Extracting Daily Infections

### Transition Matrices

```{r extract-transitions}
cat("=== Extracting E→I transitions ===\n")

transitions <- sim_results$transition %>%
  filter(from == "Exposed", to == "Infected") %>%
  arrange(sim_num, date)

cat("Dimensions:", nrow(transitions), "rows ×", ncol(transitions), "columns\n")
head(transitions, 10)

agg_by_date <- transitions %>%
  group_by(sim_num, date) %>%
  summarise(counts = sum(counts), .groups = "drop")
```

### Computing Confidence Intervals

```{r calculate-quantiles}
quantiles_df <- agg_by_date %>%
  group_by(date) %>%
  summarize(
    lower_ci = quantile(counts, 0.025),
    upper_ci = quantile(counts, 0.975),
    median   = quantile(counts, 0.5),
    mean     = mean(counts),
    .groups  = "drop"
  )

head(quantiles_df, 10)
```

## Visualization

```{r prepare-plot-data}
plot_df <- quantiles_df %>%
  mutate(
    Date           = covid_data$Date,
    observed_cases = incidence
  )
```

```{r plot-results, fig.width=11, fig.height=7}
p_with_ci <- ggplot(plot_df, aes(x = Date)) +
  geom_ribbon(
    aes(ymin = lower_ci, ymax = upper_ci),
    fill = "red", alpha = 0.3
  ) +
  geom_line(
    aes(y = median, color = "Model median (E→I)"),
    linewidth = 1.2
  ) +
  geom_line(
    aes(y = observed_cases, color = "Observed cases"),
    linewidth = 1.4
  ) +
  geom_point(
    aes(y = observed_cases, color = "Observed cases"),
    size = 2
  ) +
  scale_color_manual(
    values = c(
      "Model median (E→I)" = "red",
      "Observed cases"     = "blue"
    )
  ) +
  labs(
    title    = "Daily E→I Transitions: Observed vs Model with 95% CI",
    subtitle = paste0(
      "Population: ", format(N, big.mark = ","),
      " | β: ",     round(beta,  4),
      " | ptran: ", round(ptran, 4),
      " | crate: ", round(crate, 1),
      " | incub: ", round(incub, 1), " days",
      " | R₀: ",    round(R0, 2)
    ),
    x     = "Date",
    y     = "Daily E→I Transition Counts",
    color = "",
    fill  = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title      = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle   = element_text(hjust = 0.5)
  )

print(p_with_ci)
```

## Model Performance

```{r performance-metrics}
coverage <- plot_df %>%
  mutate(in_ci = observed_cases >= lower_ci & observed_cases <= upper_ci) %>%
  summarize(
    coverage_rate = mean(in_ci) * 100,
    n_days        = n(),
    days_in_ci    = sum(in_ci)
  )

rmse <- sqrt(mean((plot_df$median - plot_df$observed_cases)^2))
mae  <- mean(abs(plot_df$median - plot_df$observed_cases))
mape <- mean(abs((plot_df$observed_cases - plot_df$median) /
                   plot_df$observed_cases)) * 100

cat("=== Model Performance Metrics ===\n\n")
cat("95% CI Coverage:", round(coverage$coverage_rate, 1), "%\n")
cat("  (", coverage$days_in_ci, "out of", coverage$n_days, "days within CI)\n\n")
cat("Root Mean Squared Error (RMSE):", round(rmse, 2), "cases/day\n")
cat("Mean Absolute Error (MAE):",      round(mae,  2), "cases/day\n")
cat("Mean Absolute Percentage Error (MAPE):", round(mape, 1), "%\n")
```

## Discussion

### Key Results

The BiLSTM-calibrated SEIR model successfully captured COVID-19 transmission dynamics in Utah:

-   β = ptran × crate (model output): **`r round(beta, 4)`**
-   Transmission probability ptran = β / crate: **`r round(ptran, 4)`**
-   Contact rate (fixed): **`r crate`** contacts/person/day
-   Incubation period (model output): **`r round(incub, 1)`** days
-   Recovery rate (fixed): **`r round(recov, 3)`** per day
-   Basic reproduction number: **R₀ = `r round(R0, 2)`**

The model achieved **`r round(coverage$coverage_rate, 1)`%** coverage with a 95% confidence interval and mean absolute error of **`r round(mae, 2)`** cases per day.

### Advantages of This Approach

1.  **Hybrid methodology**: Combines neural network pattern recognition with mechanistic epidemiological modeling
2.  **Interpretable parameters**: Outputs β, R₀, and incubation period — key epidemiological quantities
3.  **Uncertainty quantification**: Stochastic simulations provide prediction intervals
4.  **Real-time capability**: Can be updated as new surveillance data becomes available
5.  **Explicit latent period**: SEIR model better represents COVID-19 biology

### Limitations

-   **β decomposition**: The model predicts β = ptran × crate as a composite; splitting requires fixing one component
-   **Homogeneous mixing**: Assumes uniform contact patterns across the population
-   **Constant parameters**: Does not capture time-varying transmission
-   **Closed population**: No births, deaths, or migration
-   **No spatial structure**: Treats the entire state as a single well-mixed population

### Future Directions

-   Incorporate time-varying parameters to model interventions
-   Extend to SEIRS model including waning immunity
-   Add age-structured compartments
-   Incorporate spatial structure with metapopulation models

## Conclusion

This analysis demonstrates that BiLSTM neural networks can effectively calibrate traditional compartmental epidemic models from observed incidence data. The model predicts β (= ptran × crate), R₀, and incubation period directly from the incidence time series. `ModelSEIRCONN` then uses the derived ptran and crate to generate stochastic epidemic trajectories with full uncertainty quantification.
